---
output: github_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Clinical Studies Review Tools

## Overview

`csrtools` is an R package designed to provide a workflow for meta-analysis and cumulative effect size estimation. It provides functions to compute standardised mean change (SMC), handle heterogeneity statistics, and generate high-quality visualisations.

This package is particularly suited for researchers in medicine, psychology, or any field that involves meta-analytical studies and time-series data.

## Features

- **Run meta analysis**: Functions to calculate standardised mean change and heterogeneity statistics for time-series data.

- **Cumulative effects**: Compute cumulative effect sizes and their confidence intervals with propagated uncertainty estimates over specific time points.

- **Handle missing data**: Option to impute missing data with weighted means before time-series analysis.

- **Data visualisation**: Generate forest and funnel plots, pooled SMC time-series, and cumulative SMC time-series, with options to save multi-panel layouts.


## Installation

You can load the `csrtools` package by running the following:

```{r installation, message=FALSE, warning=FALSE}
# Install devtools if not already installed
#install.packages("devtools")

# Install CSRtools from GitHub
devtools::install_github("becky-turner/csrtools")

# Once installed, load the package
library("csrtools")

```

## Load example data and other packages

Note that the example dataset `bctq_data` loaded below has been pre-processed for compatibility with the package.

```{r, message=FALSE, warning=FALSE}
# Load BCTQ example data
data(bctq_data)

# Load additional packages
library(metafor)
library(dplyr)
library(tidyr)
library(ggplot2)

```


## Example: Run standardised mean change meta-analysis

Run SMC meta analysis from baseline to 2 weeks for the open CTR group:

```{r smc1, warning=FALSE, fig.height=7, fig.width=11}
# Intervention = Open CTR
open_2weeks <- smc_tp(data=bctq_data[bctq_data$intervention_arm=="Open",], 
       outcome_mean = "mean_2weeks", 
       outcome_sd = "sd_2weeks",
       time_point = "2", time_unit = "weeks (from baseline)", 
       plot=TRUE)

```

Now run the SMC meta analysis for the endoscopic CTR group with heterogeneity statistics:


```{r smc2, message=FALSE,warning=FALSE, fig.height=7, fig.width=11}
endoscopic_2weeks <- smc_tp(data=bctq_data[bctq_data$intervention_arm=="Endoscopic",], 
       outcome_mean = "mean_2weeks", 
       outcome_sd = "sd_2weeks",
       time_point = "2", time_unit = "weeks (from baseline)", 
       plot=FALSE, heterogeneity = TRUE)

print(endoscopic_2weeks)

```


## Example: Plot pooled SMC across interventions

Here, calculate the pooled SMC across all interventions and visualise the results as a time-series.

```{r pooled smc,warning=FALSE, message=FALSE}
# Define column names and time points
outcome_means <- c("mean_1week", "mean_2weeks", "mean_3weeks", "mean_4weeks","mean_6weeks",
                   "mean_12weeks", "mean_24weeks", "mean_52weeks")

outcome_sds <- c("sd_1week", "sd_2weeks", "sd_3weeks", "sd_4weeks", "sd_6weeks",
                 "sd_12weeks", "sd_24weeks", "sd_52weeks")

time_points <- c("1", "2", "3", "4", "6", "12", "24", "52")

time_unit <- "weeks"

# Run pooled effect size analysis for all interventions

intervention <- "All"

(
  pooled_smc_all <- pooled_smc_tp(data = bctq_data,
                                      outcome_means = outcome_means,
                                      outcome_sds = outcome_sds,
                                      baseline_mean = "mean_baseline",
                                      baseline_sd = "sd_baseline",
                                      sample_size = "n_in_arm",
                                      time_points = time_points,
                                      time_unit = time_unit,
                                      intervention = intervention, plot = TRUE)
)

```


## Example: Calculate correlation coefficients

Calculate the correlation coefficients between BCTQ scores at baseline, 4 weeks and 12 weeks across all interventions.

```{r}
outcome_means <- c("mean_baseline", "mean_4weeks","mean_12weeks")

means_all <- bctq_data[,outcome_means]
correlation_table(data = means_all, save_to_doc = FALSE)

```


## Example: Estimate cumulative effects

Calculate the cumulative SMC estimates with confidence intervals for the open CTR group. Below, the `open_smc_data` subset includes only data for the Open CTR subset of patients. The dataset was curated using the `smc_tp` function above to calculate the SMC between each time-point.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
## Impute means for missing values in open subset
dat1 <- bctq_data[bctq_data$intervention_arm=="Open",]

# Define the columns to work on (columns 11 to 31)
impute_cols <- colnames(dat1)[11:31]

dat1 <- impute_missing(data = dat1, impute_cols, sample_size = "n_in_arm")

# Re-assign to open vs endoscopic subsets
open_subset <- dat1

## Open subset ----
open_1week <- smc_tp(open_subset,"mean_baseline", "sd_baseline", "mean_1week", "sd_1week",
                      sample_size = "mean_1week_sample", study_id = "author_arm", time_point="1")

open_2weeks <- smc_tp(open_subset, mean_from = "mean_1week", sd_from = "sd_1week",
                         outcome_mean = "mean_2weeks", outcome_sd = "sd_2weeks",
                         sample_size = "mean_2weeks_sample", study_id = "author_arm", time_point="2")

open_3weeks <- smc_tp(open_subset, mean_from = "mean_2weeks", sd_from = "sd_2weeks",
                         outcome_mean = "mean_3weeks", outcome_sd = "sd_3weeks",
                         sample_size = "mean_3weeks_sample", study_id = "author_arm", time_point="3")

open_4weeks <- smc_tp(open_subset, mean_from = "mean_3weeks", sd_from = "sd_3weeks",
                         outcome_mean = "mean_4weeks", outcome_sd = "sd_4weeks",
                         sample_size = "mean_4weeks_sample", study_id = "author_arm", time_point="4")

open_6weeks <- smc_tp(open_subset, mean_from = "mean_4weeks", sd_from = "sd_4weeks",
                         outcome_mean = "mean_6weeks", outcome_sd = "sd_6weeks",
                         sample_size = "mean_6weeks_sample", study_id = "author_arm", time_point="6")

open_12weeks <- smc_tp(open_subset, mean_from = "mean_6weeks", sd_from = "sd_6weeks",
                          outcome_mean = "mean_12weeks", outcome_sd = "sd_12weeks",
                          sample_size = "mean_12weeks_sample", study_id = "author_arm", time_point="12")

open_24weeks <- smc_tp(open_subset, mean_from = "mean_12weeks", sd_from = "sd_12weeks",
                          outcome_mean = "mean_24weeks", outcome_sd = "sd_24weeks",
                          sample_size = "mean_24weeks_sample", study_id = "author_arm", time_point="24")

open_52weeks <- smc_tp(open_subset, mean_from = "mean_24weeks", sd_from = "sd_24weeks",
                          outcome_mean = "mean_52weeks", outcome_sd = "sd_52weeks",
                          sample_size = "mean_52weeks_sample", study_id = "author_arm", time_point="52")

open_72weeks <- smc_tp(open_subset, mean_from = "mean_52weeks", sd_from = "sd_52weeks",
                          outcome_mean = "mean_72weeks", outcome_sd = "sd_72weeks",
                          sample_size = "mean_72weeks_sample", study_id = "author_arm", time_point="72")

open_104weeks <- smc_tp(open_subset, mean_from = "mean_72weeks", sd_from = "sd_72weeks",
                           outcome_mean = "mean_104weeks", outcome_sd = "sd_104weeks",
                           sample_size = "mean_104weeks_sample", study_id = "author_arm", time_point="104")

open_smc_data <- rbind(open_1week, open_2weeks, open_3weeks, open_4weeks, open_6weeks, open_12weeks, open_24weeks,open_52weeks, open_72weeks, open_104weeks)

# numeric time_point column
open_smc_data$time_point <- as.numeric(open_smc_data$time_point)

```


```{r}
# Sort the data by time_point to ensure proper calculation
open_smc_data <- open_smc_data[order(open_smc_data$time_point), ]

# Calculate cumulative SMC changes
open_smc_data$smc_cum <- cumsum(open_smc_data$smc)

# Calculate cumulative confidence intervals (fixed-width option)
open_smc_data <- cumulative_ci(open_smc_data, method = "wald")

# Add 0 time point
open_smc_data <- rbind(0, open_smc_data)

# Add intervention names
open_smc_data$intervention <- "Open"

# View the final dataset
head(open_smc_data)

```

## Visualise cumulative effects

Here, we use a smoothing function over the cumulative SMC estimates calculated from the step above and plot the point estimates with confidence intervals as a time-series.

```{r run csmc, warning=FALSE, message=FALSE}
smc_smooth <- smooth_csmc(open_smc_data, csmc = "smc_cum",
                          lower_ci = "lower_ci_wald", upper_ci = "upper_ci_wald",
                          intervention = "intervention",
                          span = 0.3, time_series = seq(0, 52, 0.001))


```

```{r plot csmc, echo=FALSE, message=FALSE, warning=FALSE}
# Create a subset of the data for the ribbon
smc_ribbon <- smc_smooth %>%
  filter(parameter %in% c("lower", "upper")) %>%
  pivot_wider(names_from = parameter,
              values_from = smoothed_value,
              id_cols = c(time_point, intervention)) %>%
  dplyr::rename(ymin = lower, ymax = upper)

# Create a subset of the data for the lines
smc_lines <- smc_smooth %>% dplyr::filter(parameter == "smc")

# Plotting parameters
cols <- c("Open" = "lightblue")

### PLOT SMOOTH CUMULATIVE SMCs---

ggplot() +
  # Confidence intervals
  geom_ribbon(data = smc_ribbon,
              aes(x = time_point,
                  ymin = ymin, ymax = ymax,
                  fill = intervention, group = intervention),
              alpha = 0.3) +
  # Smoothed cumulative standardised mean change
  geom_line(data = smc_lines,
            aes(x = time_point, y = smoothed_value,
                color = intervention, group = intervention),
            size = 1) +
  # Standardised mean change points
  geom_point(data = open_smc_data,
             aes(x = time_point, y = smc_cum,
                 color = intervention),
             size = 2, shape = 16) +
  # Aesthetics
  scale_color_manual(values = cols) +
  scale_fill_manual(values = cols) +
  labs(x = "Weeks",
       y = "Cumulative standardised mean change") +
  theme_minimal() +
  xlim(c(0, 52)) +
  theme(legend.position = "bottom",
        legend.title = element_blank())


```

## Documentation

For detailed documentation on all functions, check the Reference Manual or run:

```{r doc}
help(package = "csrtools")

```

## Citation

Turner R, Hartrick O (2025). csrtools: Clinical Studies Review Tools.https://doi.org/10.5281/zenodo.14841467.


## Contributing

Contributions, suggestions, and bug reports are welcome! Please use the issues page to report bugs or request features.

## Authors

- Becky Turner (maintainer)
- Olivia Hartrick

# License

This package is licensed under the MIT license. See LICENSE for details.

